COMPAS Race Analysis

### 2022-04-07 final_metrics_analysis.Rmd
adding in fairness definitions to rmd

**add brief definition of group fairness**

@@

Computed as the difference of the rate of favorable outcomes (in this case, did not recidivate) received by the unprivileged group to the privileged group. It essentially equalizes the outcomes across the protected and non-protected groups.
The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

@@

@@

```{python group fairnesss disparate impact ratio}
disp_impact_ratio = disparate_impact_ratio(y_test, y_pred, prot_attr='race', pos_label = 0)

print(f'[Baseline] The disparate impact ratio is {disp_impact_ratio: .2}')
# disparate impact ratio without y_pred attribute was 0.77
```

@@

@@

@@

This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
Fairness for this metric is between -0.1 and 0.1

@@

>>>>

With group fairness metrics, different groups should receive similar treatments or outcomes.  In this case of recidivism and race, this means that Black defendants should have similar rates of predicted recidivism as white defendants.

@@

Statistical Parity difference is computed as the difference in the rate of favorable outcomes (in this case, did not recidivate) received by the unprivileged group to the privileged group. It essentially equalizes the outcomes across the privileged and non-privileged groups.
The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$

@@

Because the statistical parity difference is negative and not within -0.1 and 0.1, the algorithm unfairly benefits white defendants.

@@

$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$

where Y is the favorable outcome of a defendant's predicted two year recidivism and D is the race of the defendant. In our case, the favorable outcome is survived, or not recidivated, which is Y = 0. In terms of race, 0 = Black and 1 = White. So to calculate the disparate impact ratio on this data, the formula is:

$\frac{P(\hat{Y}=0|D=0)}{P(\hat{Y}=0|D=1)}$

Say we have a data test set of 200 defendants, 125 Black and 75 white. Of the 125 Black defendants, 44% (55) were predicted to not recidivate. Of the 75 white defendants, 66% (50) were predicted to not recidivate.

The disparate impact ratio would therefore be

$\frac{0.44}{0.66} = 0.667$

Because this value is less than 1, it benefits white defendants. Additionally, because it is below 0.8, it is considered "unfair."

```{python group fairnesss disparate impact ratio}
disp_impact_ratio = disparate_impact_ratio(y_test, y_pred, prot_attr='race', priv_group = 1, pos_label = 0)

print(f'[Baseline] The disparate impact ratio is {disp_impact_ratio: .2}')
```

Since the actual disparate impact ratio for our test data is 0.47 and below 0.8, the algorithm unfairly benefits white defendants i.e. disproportionately predicts them to not recidivate.

@@

This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 is considered fair, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both black and white defendants.

$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$

@@

Since the average odds difference is below 0 by a substantial amount, white defendants are unfairly benefitted by the algorithm.

@@

This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
Fairness for this metric is between -0.1 and 0.1.

$TPR_{D = Unprivileged} - TPR_{D = Privileged}$

@@

Since the equal opportunity difference is below 0 and beyond -0.1, the algorithm unfairly benefits white defendants.
